Reproducible Builds for\\ Computational Research Papers

### 2020-09-14 ms.tex
[make test]

We propose a template for writing computational research papers with \LaTeX and verifying the its reproducibility using the concept of \textit{reproducible builds} (RBs) in a local or remote computational environment.
We provide an open source implementation of the template\footnote{\url{https://github.com/pbizopoulos/cookiecutter-reproducible-builds-for-computational-research-papers}} and use it to write this paper\footnote{\url{https://arxiv.org/abs/2005.12660}} and verify its reproducibility\footnote{\url{https://github.com/pbizopoulos/reproducible-builds-for-computational-research-papers}}.
We use Python~\cite{van2007python} for producing the results, however other scientific-oriented programming languages can be used.

@@

The proposed template can be applied on \textit{papers} written in \LaTeX~\cite{lamport1994latex} and the \textit{results code} are executed within a Docker container.
\LaTeX\ is a typesetting system for publishing high quality research papers, and Docker~\cite{merkel2014docker} is a container system that allows portable execution of a codebase between different operating systems/environments and has shown great promise in reproducible research~\cite{boettiger2015introduction}.
The corresponding Dockerfile can be provided by the \textit{author} or generated by tools such as containerit~\cite{nust2019containerit}.

A specific file hierarchy (shown in Fig.~\ref{fig:filehierarchyworkflow}) is used in this paper however this is optional, as the template only needs to know the directories of the main files of the \LaTeX\ and \textit{results code} during its creation by the template.
The template fits in the research writing and reproducibility verification in the following way:
\begin{enumerate}
	\item the \textit{author} edits the \textit{code} in the local repository,
	\item the \textit{author} may execute \textit{make} during development and experimentation,
	\item the \textit{author} may verify the reproducibility of the \textit{paper} locally by executing \textit{make test} (or remotely by commiting with a message which contains \textit{[make test]}) which triggers the following:
		\begin{enumerate}
			\item the \textit{results code} is executed,
			\item the \textit{results} along with the \textit{\LaTeX\ code} are used to compile a pdf file,
			\item the previous is executed twice,
			\item the \textit{paper} is verifed as reproducible if and only if the pdf files are identical.
		\end{enumerate}
	\item the \textit{reviewer} may verify the reproducibility of the paper on a local or remote machine.
\end{enumerate}

Docker usage in RBs provides portability of the environment of the research while the VM provides trust in the procedure (in contrast with verifying locally, thus having low credibility in reporting reproducibility).
The reason that two local/remote compilations are needed is to enable comparison between different builds of the same source code and any non-deterministic stochasticity (such as not applying a specific seed to random generators) would produce different pdfs, thus making the \textit{paper} non-reproducible.

@@

Other potential uses of the RBs template include:
\begin{itemize}
	\item Preprint server (such as arXiv) \textit{paper} reproducibility verification.
		The difference with the remote reproducibility verification is the presence of an additional VM that checkouts the \textit{\LaTeX\ code} and generates a pdf which is then compared with the one generated from \textit{code}.
	\item Regression testing and debugging, to ensure that changes to \textit{code} do not alter the results.
\end{itemize}

\subsection{Technical implementation}
An open source implementation of the RBs template is available in which the remote reproducibility verification is done using Github Actions.
The builder VMs use the latest version of Ubuntu and pull a docker image for building \LaTeX\ from the Github Container Registry.
We demonstrate its use by writing and verifying the reproducibility of this paper and its arXiv version.

Regarding research that is computational costly, the template by default allows the execution and reproducibility verification of a computationally lighter version of the \textit{paper} and provides an argument for executing the whole experiment.
This also helps in faster iterations during research writing and experimentation.
An example of a computationally lighter version in the field of neural networks is setting the number of epochs and number of training samples to a low value.

A common culprit against reproducibility in \LaTeX\ is the time-date metadata of pdf output.
These can be disabled using the following into the \textit{\LaTeX\ code} after the preamble or as extra arguments during \LaTeX\ compilation (as done in the RB Makefile):
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ pdf reproducibility commands.}, captionpos=b]
\pdfinfoomitdate=1
\pdfsuppressptexinfo=-1
\pdftrailerid{}
\end{lstlisting}

@@

The RB template takes advantage of the automatic generation of intermediate files of the `make' program.
The \textit{author} only needs to edit (or use the `touch' UNIX utility to force) a file from \textit{results code} or \textit{\LaTeX\ code} and the corresponding compilation is automatically triggered.

\section{Example Use Case of RBs}
This sections provides a use case of RBs and also serves as a manual for creating reproducible \textit{papers} written in \LaTeX\ with Python as the \textit{results code} programming language.
For demonstration purposes we train and test a VGG11 neural network~\cite{simonyan2014very} with batch normalization with PyTorch~\cite{paszke2019pytorch}, on the following datasets:
\begin{itemize}
	\item MNIST~\cite{lecun2010mnist}
	\item FashionMNIST~\cite{xiao2017fashion}
	\item KMNIST~\cite{clanuwat2018deep}
	\item QMNIST~\cite{yadav2019cold}
\end{itemize}

@@

For example the values of the following variables are not referred in the main \textit{.tex} file but they are read by an intermediate \textit{.tex} file created by the \textit{results code}:
\begin{itemize}
	\item the learning rate is $\DTLfetch{keys_values}{key}{lr}{value}$,
	\item the batch size is $\DTLfetch{keys_values}{key}{batch_size}{value}$ and
\end{itemize}

Regarding the \textit{results code} for Python, random seeds need to be set to a specific value such as:
\begin{lstlisting}[language=python, style=lststyle, caption={Python reproducibility commands for some popular libraries.}, captionpos=b]
# for build-in random module
random.seed(0)
# for numpy
np.random.seed(0)
# for pytorch
torch.backends.cudnn.benchmark = False
torch.backends.cudnn.deterministic = True
torch.cuda.manual_seed_all(0)
torch.manual_seed(0)
# for Tensorflow
tf.random.set_seed(0)
\end{lstlisting}

Using the random seeds and \textit{datatool} we could have deterministic stochasticity that reproduces figures, tables and variables.
For example Fig.~\ref{fig:image} depicts the train and validation loss:
\begin{figure*}[!t]
	\subfloat{\includegraphics[width=0.25\textwidth]{results/MNIST-loss}}
	\subfloat{\includegraphics[width=0.25\textwidth]{results/FashionMNIST-loss}}
	\subfloat{\includegraphics[width=0.25\textwidth]{results/KMNIST-loss}}
	\subfloat{\includegraphics[width=0.25\textwidth]{results/QMNIST-loss}}
	\caption{Train validation lossses.
	Image example created from \textit{results code}.}
	\label{fig:image}
\end{figure*}

Another useful function for Python is \textit{pandas.DataFrame.to\_latex} which automatically converts a dataframe table to a \LaTeX\ table (as shown in Table~\ref{table:table}.

>>>>

We propose a template for writing computational research papers with \LaTeX\ and verifying the its reproducibility using the concept of \textit{reproducible builds} (RBs) in a local or remote computational environment.
We provide an open source implementation of the template\footnote{\url{https://github.com/pbizopoulos/cookiecutter-reproducible-builds-for-computational-research-papers}} and use it to write this paper\footnote{\url{https://arxiv.org/abs/2005.12660}} and verify its reproducibility\footnote{\url{https://github.com/pbizopoulos/reproducible-builds-for-computational-research-papers}}.
We use Python~\cite{van2007python} for producing the results, however other scientific-oriented programming languages can be used.

@@

The proposed template can be applied on \textit{papers} written in \LaTeX~\cite{lamport1994latex} and the \textit{results code} are executed within a Docker container.
\LaTeX\ is a typesetting system for publishing high quality research papers, and Docker~\cite{merkel2014docker} is a container system that allows portable execution of a codebase between different operating systems/environments and has shown great promise in reproducible research~\cite{boettiger2015introduction}.
The corresponding Dockerfile can be provided by the \textit{author} or generated by tools such as repo2docker~\cite{forde2018reproducible}.

A specific file hierarchy is shown in Fig.~\ref{fig:filehierarchyworkflow} however this is optional, as the template only needs to know the directories of the main files of the \LaTeX\ and \textit{results code} during its creation.
The template fits in the research writing and reproducibility verification in the following way:
\begin{enumerate}
	\item the \textit{author} edits the \textit{code} in the local repository,
	\item the \textit{author} may execute \textit{make} during development and experimentation,
	\item the \textit{author} may verify the reproducibility of the \textit{paper} locally by executing \textit{make test} (or remotely by commiting with a message which contains \textit{[make test]}) which triggers the following:
		\begin{enumerate}
			\item the \textit{results code} is executed,
			\item the \textit{results} along with the \textit{\LaTeX\ code} are used to compile a pdf file,
			\item the previous is executed twice,
			\item the \textit{paper} is verifed as reproducible if and only if the pdf files are identical.
		\end{enumerate}
	\item the \textit{reviewer} may verify the reproducibility of the paper locally (by cloning the repo) or remotely (by inspecting the result of \textit{[make test]}).
\end{enumerate}

The reason that two local/remote compilations are needed is to enable comparison between different builds of the same source code and any non-deterministic stochasticity (such as not applying a specific seed to random generators) would produce different pdfs, thus making the \textit{paper} non-reproducible.
Moreover Docker usage in RBs provides portability of the environment of the research while third-party VMs are used when the verification needs to be shared (in contrast with verifying locally, thus having low credibility in reporting reproducibility).

@@

Potential uses of the RBs template include:
\begin{itemize}
	\item preprint server (e.g.\ arXiv) \textit{paper} reproducibility verification and
	\item regression testing and debugging, to ensure that changes to \textit{code} do not alter the results.
\end{itemize}

\subsection{Technical implementation}
We provide an open source implementation of the RBs template in which the remote reproducibility verification is done using Github Actions.
The builder VM uses the latest version of Ubuntu and pulls a docker image for building \LaTeX\ from the Github Container Registry.
We demonstrate its use by writing and verifying the reproducibility of this paper.

@@

Regarding research that is computational costly, the template allows the execution and reproducibility verification of a computationally lighter version of the \textit{paper} by default.
This also helps in faster iterations during research writing and experimentation.
An example of computationally lighter versions in the field of neural networks is setting the number of epochs and number of training samples to a low value.
The execution of the full experiment is done by providing an additional argument.

The RB template takes advantage of the automatic generation of intermediate files of the `make' program.
The \textit{author} only needs to edit (or use the `touch' UNIX utility to force) a file from \textit{results code} or \textit{\LaTeX\ code} and the corresponding compilation is automatically triggered with `make'.

\section{Example Use Case of RBs}
This sections provides a use case of RBs and also serves as a manual for creating reproducible \textit{papers} written in \LaTeX\ with Python as the \textit{results code} programming language.
For demonstration purposes we train and test a VGG11 neural network~\cite{simonyan2014very} with batch normalization in PyTorch~\cite{paszke2019pytorch}, on MNIST~\cite{lecun2010mnist}, FashionMNIST~\cite{xiao2017fashion}, KMNIST~\cite{clanuwat2018deep} and QMNIST~\cite{yadav2019cold}.

Regarding \LaTeX\ a common culprit against reproducibility is the time-date metadata embedded in the pdf output.
These can be disabled using the following commands into the \textit{\LaTeX\ code} or as extra arguments during \LaTeX\ compilation (as done in the RB Makefile):
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ pdf reproducibility commands.}, captionpos=b]
\pdfinfoomitdate=1
\pdfsuppressptexinfo=-1
\pdftrailerid{}
\end{lstlisting}

@@

For example the values of the following variables are not referred in the main \textit{.tex} file but they are read by an intermediate \textit{.tex} file created by the \textit{results code}:
\begin{itemize}
	\item the number of epochs is $\DTLfetch{keys_values}{key}{num_epochs}{value}$,
	\item the batch size is $\DTLfetch{keys_values}{key}{batch_size}{value}$ and
	\item the learning rate for SGD is $\DTLfetch{keys_values}{key}{lr}{value}$.
\end{itemize}

Regarding the \textit{results code} for Python, random seeds need to be set to a specific value such as:
\begin{lstlisting}[language=python, style=lststyle, caption={Python reproducibility commands for some popular libraries.}, captionpos=b]
random.seed(0) # for build-in random module
np.random.seed(0) # for numpy
tf.random.set_seed(0) # for Tensorflow
# for pytorch
torch.backends.cudnn.benchmark = False
torch.backends.cudnn.deterministic = True
torch.cuda.manual_seed_all(0)
torch.manual_seed(0)
\end{lstlisting}

Using the random seeds and \textit{datatool} we could have deterministic stochasticity that reproduces figures, tables and variables.
For example Fig.~\ref{fig:image} depicts the train and validation loss:
\begin{figure*}[!t]
	\subfloat{\includegraphics[width=0.25\textwidth]{results/MNIST-loss}}
	\subfloat{\includegraphics[width=0.25\textwidth]{results/FashionMNIST-loss}}
	\\
	\subfloat{\includegraphics[width=0.25\textwidth]{results/KMNIST-loss}}
	\subfloat{\includegraphics[width=0.25\textwidth]{results/QMNIST-loss}}
	\caption{Train validation losses.
	Image example created from \textit{results code}.}
	\label{fig:image}
\end{figure*}

Another useful function for Python is \textit{pandas.DataFrame.to\_latex} which automatically converts a dataframe table to a \LaTeX\ table (as shown in Table~\ref{table:table}).
